A decision support tool for the Home Health Care Routing and Scheduling Problem

### 2021-09-09 tex/Paper/main.tex
main file for paper.

\documentclass[a4paper,11pt]{elsarticle}

\input{includes/preamble.tex}

\biboptions{authoryear}

\journal{European Journal of Operational Research}

\begin{document}

@@

%% Title, authors and addresses

@@

\author[sbs]{Carlos Lamas-Fernandez$^{*,}$}
\author[exeter]{Thomas Monks}
\author[sbs]{Antonio Martinez-Sykora}

\address[sbs]{Southampton Business School, University of Southampton, Southampton SO17 1BJ, UK}
\address[exeter]{University of Exeter Medical School, University of Exeter, Exeter, UK}

\begin{abstract}
%% Text of abstract
The Home Health Care Routing and Scheduling Problem (HHCRSP) consists in assigning routes and schedules to nurses that deliver health care services to patients at their homes. A varied range of hard and soft constraints such as time windows, synchronisation of nurses, skill mix or patient and nurse preferences make this problem particularly difficult to solve manually. However, this is a challenge faced daily by many community nursing teams.

@@

\begin{keyword}
OR in health services \sep Metaheuristics \sep Routing \sep Scheduling \sep GRASP
%% keywords here, in the form: keyword \sep keyword

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

@@

{\footnotesize
\noindent $^*$ Corresponding author\\
\noindent E-mail addresses: \text{C.Lamas-Fernandez@soton.ac.uk} (C. Lamas-Fernandez), \text{T.M.W.Monks@exeter.ac.uk} (T. Monks), \text{A.Martinez-Sykora@soton.ac.uk} (A. Martinez-Sykora)
}

\section{Introduction}\label{seq:intro}

Home health care refers to a form of care where a health professional visits a patient in their own home to perform a health-related task \citep{Cisse2017}. In the European context, with an ever-ageing population, delivering health care at home is increasingly seen as a necessary alternative to traditional forms of care \citep{tarricone2008home}.
This view is shared by many health providers around the globe as managing long term conditions, frail patients and end-of-life care among others might be done more efficiently in the patient's own home. For example, in the OECD countries, the care at home for patients with long term conditions increased by 6\% \citep{OECDHealthataGlance2019} and in the UK the National Health Service (NHS) stated in their long term plan that in the future more care should be delivered to patients in their own homes \cite{england2019nhs}.

A key role in the health service to deliver this care are District Nurses (DN), specialised nursing teams that travel to patient's homes and perform a variety of tasks, ranging from insulin injections to dressing wounds. Nevertheless, DN numbers have fallen by 44\% \cite{NHSProviders2018} between 2010 and 2018 in the NHS. On the other hand, while there is no official data on demand, the workforce perception is that demand has risen significantly, putting the service under pressure \citep{robertson2017understanding}. There is, therefore, a strong need for efficiency within the community nursing service, to better utilise the available resources.

@@

In an effort to automate and improve this process, we model the planning of home health care visits as a rich vehicle routing problem, with constraints and objectives drawn from our interaction with community nursing teams. This problem is often known in the literature as Home Health Care Routing and Scheduling Problem (HHCRSP). In our case, the schedule takes place over a single day, as it is the preferred approach in practice. This is motivated by the last-minute changes due to staff availability, patient next-day referrals and the nature of continuous review of care, which often prompts to scheduling the next visit right after reviewing their condition on the previous visit. A formal description of the problem is provided in Section~\ref{seq:model description}.

We propose a Decision Support Tool (DST) that can produce high quality daily routes and schedules for a DN team, according to a flexible objective function. The tool requires as input a patient requirement list and the available staff and then runs an optimisation algorithm based on a weighted objective function. The optimisation is based on a Greedy Randomised Adaptive Search Procedure (GRASP) with Path Relinking, which is detailed in Section~\ref{seq:DSS}.

The main contribution of this article is to present a DST for the HHCRSP that is designed to be used in real community nursing settings. It tackles the most common constraints found in literature (e.g. time windows or skills) but also others often found in practice (e.g. multiple depots or workload balance). We provide extensive computational results (Section~\ref{seq:Computational Experiments}) that show that our algorithm is competitive with state-of-the-art approaches and is flexible to allow real-world constraints and objectives; but also provide insight on the algorithm components that help achieve high quality routes and schedules.

\section{Literature review}
The generalised growth in demand and availability of home health care services of recent years has resulted in a growth in academic interest in the related scheduling problem. While the problem might be addressed by drawing from the extensive vehicle routing problem (VRP) literature (see, for example, \cite{Drexl2012} for a comprehensive review VRPs and synchronisation constraints), HHCRSP has developed its own body of literature. In particular, there are two recent review articles, \cite{Cisse2017} and \cite{Fikar2017}. We refer the reader to them for a broader understanding of what it is meant by HHCRSP and different solving approaches. Home Health Care and Home Care services across the world operate in different ways and, as such, the related scheduling problems have different objectives and constraints. For example, it might be possible to speed up the visits to some patients \citep{Mosquera2018}, to allow nurses to share a minibus for part of their route, while walking the rest of it \citep{Fikar2015}, for nurses to use public transport \citep{Rest2016}, or to allow time preferences for patients \citep{Decerle2018}. % This variety makes difficult comparing approaches and translating the findings from one application to another.

@@

% Exact methods
Several articles propose exact methods. These are usually based on a Mixed Integer Linear Program (MILP) formulation that is solved with the help of a commercial solver \citep{Bredstrom2008,Bachouch2011, Mankowska2014, AitHaddadene2016}. Typically, these formulations are three index formulations that involve two types of variables: $x_{ijk}$ binary variables that indicate if carer $k$ traverses the arc $i$ and $j$; and $t_{ik}$ real variables that determine at what time carer $k$ starts performing job $i$. These formulations are limited in practice and can only solve to optimality small problems (\eg 45 patients in \cite{AitHaddadene2016} or 10 in \cite{Mankowska2014}). In order to address this, some researchers propose to decompose the problem in a column generation scheme. For example, \cite{Rasmussen2012} tackles the HHCRSP problem in Denmark. They propose a branch-and-price scheme where generalised synchronisation constraints are relaxed and dealt with in the branching. % \correction{Only Rasmussen?}
 % \correction{Cite relevant papers here}. The ILP formulations work well for small problems, but are often not practical to solve for problems with a significant number of patients. %around 50

Often, in order to tackle larger instances, the MILPs are solved in a restricted manner and, while they do not guarantee optimality, they might provide feasible solutions suitable for day to day planning. This is often referred to as a matheuristic algorithm and can be done by reducing the size of the original problem by fixing some aspects and solving it again. One example of this technique is given in \cite{Rasmussen2012} that also proposes clustering strategies to reduce the size of the problem, which allows them to solve problems with up to $150$ patients, although no longer proving optimality. Another is \cite{Bredstrom2008}, where they propose a heuristic that iteratively allows or forbids assignments between patients and carers in order to generate a series of restricted MILPs that can be solved in a short time.

Another approach, which is the one we use in this article, is to propose a metaheuristic algorithm independent of a mathematical solver. \cite{Mankowska2014} and \cite{AitHaddadene2016} propose an Adaptive Large Neighbourhood search and a hybrid GRASP combined with Iterated Local Search respectively. One advantage of the use of heuristic algorithms is that they do not require availability of a (often commercial) MILP solver, which is unlikely to be available to health care professionals.

Finally, we look at the literature that discusses the software implementation of the solution approaches in practice. One of the earliest works is due to \cite{Begur1997}, who developed a decision support system based on the Clarke and Wright algorithm \citep{Clarke1964} for vehicle routing and then relied on integration with a Geographic Information System (GIS), an interactive user interface and a human planner to alter the routes as necessary. The Laps Care system \citep{Eveborn2006, Eveborn2009a} is a Swedish software for the HHCRSP, based on a repeated matching algorithm has been shown to be successful in saving planning time, minimising travel times and even reducing the number of missed visits. The authors estimated annual savings in 2009 of up to €120 M across teams in Norway, Finland and Sweden \citep{Eveborn2009a}. Another software is Parpap \citep{BERTELS20062866}, which proposes various approaches combining constraint programming and metaheuristics. A common claim in all these implementation-oriented works is that the software needs to be flexible enough to allow for real-world constraints to be implemented and offers interaction with mapping software.

%%%%%%%%%%%%%%%%% MODEL FOR HHCRSP %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A model for health home care services}\label{seq:model description}

In this section we describe the notation and give a formal description of the HHCRSP problem tackled by our DST.

We consider a list of nurses $N = \{1,\dots,n\}$ that are available for the day. Each nurse $i$ might have a different starting time, $st(i)$, and end time, $en(i)$, of their working day. During the planning horizon, which is one day in our case, the nurses need to perform a number of jobs $J = \{1,\dots,m\}$, a subset of which, $DS \subset J$ are double services, \ie, they require two nurses to be present to be performed. Note that jobs refer to individual tasks that a nurse needs to perform and are not necessarily identified with unique patients or locations, meaning that the same patient might have more than one job on the list.

@@

\subsection{Skill mix}
Some jobs might require a specialised nurse to be performed, and nurses might only perform jobs that they are qualified to do. To model this, we define a function $S: N\times J \rightarrow \{0,1\}$, that takes value $1$ if nurse $i$ is skilled to perform job $j$, $S_1(i,j) = 1$ and $0$ otherwise. 
Naturally, for double service jobs, no single nurse is skilled to perform them $S_1(i,j) = 0, \forall i \in N, \forall j \in DS$. To account for this, we define a second function $S_2: N\times N\times J \rightarrow \{0,1\}$, that takes value $S_2(i_1,i_2,j) = 1$ if the combination of nurses $i_1$ and $i_2$ is sufficiently skilled to perform job $j$ and $0$ otherwise.

@@

A relevant modelling consideration relates to the necessary skill mix for simultaneous double services or, in other words, how the function $S_2$ is created in practice. We distinguish the following cases:
\begin{itemize}
    \item \textbf{Duplicated}. In this case it is required for all staff members visiting the patient to have all necessary skills to perform the service.
    \item \textbf{Shared}. This is the case when it is sufficient that the combination of skills between the staff members present is enough to perform the service. In particular, it could happen that only one staff member is qualified for all services and the other for none. An example of this would be visits that require a second person to provide support and ensure safety, for example with patients that have a record of behaving inappropriately towards staff.
    \item \textbf{Strictly shared}. In this case each member of the service does not need to be qualified for all services but should be attending the visit to perform at least one service. This is the approach used in \cite{Mankowska2014} and \cite{AitHaddadene2016}.
    \item \textbf{Shared-Duplicated}. This case is analogous to \emph{Shared}, but it allows some skills to be required by more than one of the staff members, hence it generalises the previous cases.
\end{itemize}

In our DST, the default approach is \emph{Shared-Duplicated}, as it allows for the flexibility that is observed in real operations. The data is introduced as a list of skills and, if some skills need to be present in more than one staff member, they can simply appear duplicated on the list. We have, however, also implemented the Strictly Shared approach, in order to solve the instances from the literature.

% \subsubsection{Hierarchical skills and downgrading}
It is not uncommon in the literature to model the skills of the workforce as hierarchical, \ie each nurse is assigned a skill level and it is assumed that they possess all the skills that lower levels have, and some more. The model might then introduce a penalty or limit for high skilled nurses performing too many low skilled jobs, known as downgrading \cite[see, \eg][]{Fikar2015}. In the DN teams we observed, skills were not hierarchical, neither were any such limits in place. Nevertheless, nurses pointed out that high skilled nurses were typically assigned only the high skilled jobs, as they might need to perform other jobs (\eg attending meetings or other administrative tasks). While this is not something that we consider explicitly, a similar objective may be achieved through setting negative preferences between high skilled nurses and low skilled jobs.

\subsection{Preferences}
It is not uncommon in practice that district nurses do their planning using preferences. These might arise from multiple reasons, such as a preserving continuity of care, gender preference, patients and nurses speaking the same language, previous conflicts between a patient and a staff member, etc. We model this with a preference function, $P: N\times J \rightarrow \mathbb{R}$ such that $P(i,j)$ represents the preference score of assigning job $i$ to staff member $j$. The function $P$ should assign a higher (positive) score to preferred patient-nurse assignments and a lower (negative) score to assignments that are better avoided. We give an indication on how to choose the specific values of the function in Section~\ref{seq:adequate_weights}.

\subsection{Travel times}
To describe travel times, for each nurse we define $N$ functions $T_i: J\cup\{k^i_0\} \times J\cup\{k^i_0\} \rightarrow \mathbb{R}$, $i \in N$, where $T_i(j_1, j_2)$ indicates the time it takes nurse $i$ to travel from job $j_1$ to job $j_2$, where $k^i_0$ refers to the start and end location of the nurse $i$. Each nurse might start from a different depot (typically, their own home), a fact that is accounted for in these functions. Note that with this definition we allow to have different travel times between jobs for different nurses, and this can be used to model, for example, different means of transport. In our observations, however, all nurses used their own vehicles, so we assume that $T_i$ returns driving times for all nurses and, in particular that:
\begin{equation}
    T_{i_1}(j_1, j_2) = T_{i_2}(j_1, j_2),\ \forall i_1, i_2 \in N,\ \forall j_1, j_2 \in J
\end{equation}

Note that this function can be substituted by a more sophisticated one without any changes to the algorithms described here.

@@

\subsection{Temporal constraints}\label{seq:sub temporal constraints}
Each job $j$ has a service time $s_j$ and a time window that is defined by its earliest start time, $e_j$ and the latest start time $l_j$.

It is a hard constraint that jobs cannot start before their earliest start time,
\begin{align}
    t_j \geq e_j, & \forall j \in J
\end{align}

However, they might start late. The tardiness with which a job $j$ starts, $z_j$ can be simply calculated by:
\begin{equation}
    z_j = \max(0, t_j - l_j)
\end{equation}

Note that tardiness here refers only to the difference between the job start and its original time window. Jobs that have dependencies or require two nurses might be further restricted temporally. This might happen in practice, for example, in situations where a certain medication must be administered a few hours after the first dose. If two jobs $j_1$ and $j_2$ are dependent, this is modelled by imposing a second time window on $j_2$, that depends on the start time of $j_1$, $[e^{(j_1)}_{j_2} + t_{j_1}, l^{(j_1)}_{j_2} + t_{j_1}]$, then, the same time window constraints hold:
\begin{equation}
    t_{j_2} \geq e^{(j_1)}_{j_2} + t_{j_1}, % & \forall j \in J
\end{equation}
\begin{equation}
    t_{j_2} \leq l^{(j_1)}_{j_2} + t_{j_1}, % & \forall j \in J
\end{equation}
Similarly, $j_2$ will impose a time window on $j_1$, and in a valid instance of the problem it must hold that:
\begin{equation}
    e^{(j_1)}_{j_2} = -e^{(j_2)}_{j_1}
\end{equation}
and
\begin{equation}
    l^{(j_1)}_{j_2} = -l^{(j_2)}_{j_1}
\end{equation}

Then, we can define the \emph{gap tardiness} of job $j_2$, $\tilde{z}_{j_2}$ as:
\begin{equation}
    \tilde{z}_{j_2} = \max(0, t_{j_2} - (l^{(j_1)}_{j_2} + t_{j_1})]
\end{equation}

Since some jobs have time windows or might need to be synchronised, waiting times might occur. We denote by $w_{ij}$ the waiting time of nurse $i$ at job $j$. Note that $w_{ij} = 0$ if $j \notin r_i$ and, otherwise, $w_{ij}$ represents the difference between $t_j$ and the arrival time of nurse $i$ to job $j$.% , denoted by $a(i,j)$.

The arrival time $a(i,j)$ of the nurse $i$ at the job $j \in r_i$ is calculated as follows:

\begin{equation}
    a(i,j) = st(i) + \sum_{h \in 0}^{j' - 1} w_{ik^i_h} + s_{k^i_h} + T_i(k^i_h, k^i_{h + 1})
\end{equation}
Where $j'$ denotes the position of job $j$ in the route $r_i$, \ie $j = k^i_{j'} \in r_i$. Note that there is not waiting or service times at the depot, \ie $w_{ik^i_0} = 0$ and $s_{k^i_0} = 0 \ \forall i \in \mathcal{N}$.

In practice, waiting times might occur for a variety of reasons. In order to maintain simplicity in our model, we only consider three reasons for which a nurse $i$ might have waiting time $w_{ij} \geq 0$ at the job $j \in r_i$:
\begin{itemize}
    \item Respecting a time window: 
    \begin{align}
        w_{ij} \geq e_i - a(i,j) & \forall i \in N, \forall j \in J
    \end{align}
    \item Waiting for a second nurse $i'$ to arrive for a double service
    \begin{align}
        w_{i j} \geq t_j - a(i',j) & \forall j \in DS,\ such\ that\ j \in r_{i} \cap r_{i'}
    \end{align}
    \item Waiting for the minimum gap of a synchronised service. If the job $j$ depends on $j'$, then:
    \begin{equation}\label{eq:waiting_for_synchronised}
        w_{ij} \geq e^{(j')}_{j} - a(i,j)
    \end{equation}
\end{itemize}

Further, we assume that waiting times are never larger than strictly necessary. To achieve this, we consider that the starting times of jobs are fixed to their earliest possible time, based  in the order in which they appear in the solution. Let $j_1, j_2 \subset J, j_1 \neq j_2$, be two jobs that have a dependency, and such that $j_1 \in r_{i_1}$ and $j_2 \in r_{i_2}$ and nurse $i_1$ appears earlier than nurse $i_2$ in the vector of nurse order, $\mathcal{O}$. Then, we enforce that:
%  $i_1 < i_2$, \ie $j_1$ appears earlier in the vector $\mathcal{O}$ than $j_2$. Then, we enforce that:
\begin{equation}\label{eq:earliest_possible_start_assumption}
    w_{i_1 j_1} = \max \{ a(i,j_1), e_{j_1} \}
\end{equation}

and

\begin{equation}\label{eq:earliest_possible_start_assumption2}
    t_{j_1} = \max \{ a(i,j_1), e_{j_1} \}
\end{equation}

Note that this implies that the order in which nurse routes appear in a solution may have an impact in the solution quality. To mitigate this effect, in Section~\ref{seq:DSS} we describe strategies to search different orderings of nurses when solving the problem.

>>>>

\documentclass[a4paper,11pt,authoryear]{elsarticle}
\input{includes/preamble.tex}

\begin{comment}
\begin{itemize}
    \item Change all terms `nurse(s)' to `carer(s)' and `patient(s)' to `client(s)'
\end{itemize}
\end{comment}
\begin{document}

@@

@@

\author[1]{Carlos Lamas-Fernandez\corref{cor1}}
\ead{C.Lamas-Fernandez@soton.ac.uk}
\cortext[cor1]{Corresponding author}
\address[1]{Southampton Business School, University of Southampton, Southampton SO17 1BJ, UK}

\author[2]{Thomas Monks}
\ead{T.M.W.Monks@exeter.ac.uk}
\address[2]{University of Exeter Medical School, University of Exeter, Exeter, UK}
\author[1]{Antonio Martinez-Sykora}
\ead{A.Martinez-Sykora@soton.ac.uk}

\begin{abstract}
The Home Health Care Routing and Scheduling Problem (HHCRSP) consists in assigning routes and schedules to nurses that deliver health care services to patients at their homes. A varied range of hard and soft constraints such as time windows, synchronisation of nurses, skill mix or patient and nurse preferences make this problem particularly difficult to solve manually. However, this is a challenge faced daily by many community nursing teams.

@@

\begin{keyword}
OR in health services \sep Metaheuristics \sep Routing \sep Scheduling \sep GRASP
\end{keyword}

\newpageafter{abstract}

@@

%%
%% Start line numbering here if you want
%%
%% \linenumbers

%----------------------------------------------------------------------------------%
\section{Introduction}
\label{sec:intro}
\noindent Home health care refers to a form of care where a health professional visits a patient in their own home to perform a health-related task \citep{cisse2017}. In the European context, with an ever-ageing population, delivering health care at home is increasingly seen as a necessary alternative to traditional forms of care \citep{tarricone2008}.
This view is shared by many health providers around the globe as managing long term conditions, frail patients and end-of-life care among others might be done more efficiently in the patient's own home. For example, in the OECD countries, the care at home for patients with long term conditions increased by 6\% \citep{oecd2019} and in the UK the National Health Service (NHS) stated in their long term plan that in the future more care should be delivered to patients in their own homes \citep{nhs2019}.

A key role in the health service to deliver this care are District Nurses (DN), specialised nursing teams that travel to patient's homes and perform a variety of tasks, ranging from insulin injections to dressing wounds. Nevertheless, DN numbers have fallen by 44\% \citep{nhs2018} between 2010 and 2018 in the NHS. On the other hand, while there is no official data on demand, the workforce perception is that demand has risen significantly, putting the service under pressure \citep{robertson2017}. There is, therefore, a strong need for efficiency within the community nursing service, to better utilise the available resources.

@@

In an effort to automate and improve this process, we model the planning of home health care visits as a rich vehicle routing problem, with constraints and objectives drawn from our interaction with community nursing teams. This problem is often known in the literature as Home Health Care Routing and Scheduling Problem (HHCRSP). In our case, the schedule takes place over a single day, as it is the preferred approach in practice. This is motivated by the last-minute changes due to staff availability, patient next-day referrals and the nature of continuous review of care, which often prompts to scheduling the next visit right after reviewing their condition on the previous visit. A formal description of the problem is provided in Section~\ref{sec:model}.

We propose a Decision Support Tool (DST) that can produce high quality daily routes and schedules for a DN team, according to a flexible objective function. The tool requires as input a patient requirement list and the available staff and then runs an optimisation algorithm based on a weighted objective function. The optimisation is based on a Greedy Randomised Adaptive Search Procedure (GRASP) with Path Relinking, which is detailed in Section~\ref{sec:dst}.

The main contribution of this article is to present a DST for the HHCRSP that is designed to be used in real community nursing settings. It tackles the most common constraints found in literature (e.g. time windows or skills) but also others often found in practice (e.g. multiple depots or workload balance). We provide extensive computational results (Section~\ref{sec:computationalexperiments}) that show that our algorithm is competitive with state-of-the-art approaches and is flexible to allow real-world constraints and objectives; but also provide insight on the algorithm components that help achieve high quality routes and schedules.

%----------------------------------------------------------------------------------%
\section{Literature review}
\label{sec:litreview}
\noindent The generalised growth in demand and availability of home health care services of recent years has resulted in a growth in academic interest in the related scheduling problem. While the problem might be addressed by drawing from the extensive vehicle routing problem (VRP) literature (see, for example, \cite{drexl2012} for a comprehensive review of VRPs and synchronisation constraints), HHCRSP has developed its own body of literature. In particular, there are two recent review articles, \cite{cisse2017} and \cite{fikar2017}. We refer the reader to them for a broader understanding of what is meant by HHCRSP and different solving approaches. Home Health Care and Home Care services across the world operate in different ways and, as such, the related scheduling problems have different objectives and constraints. For example, it might be possible to speed up the visits to some patients \citep{mosquera2019}, to allow nurses to share a minibus for part of their route, while walking the rest of it \citep{fikar2015}, for nurses to use public transport \citep{rest2016}, or to allow time preferences for patients \citep{decerle2018}. % This variety makes difficult comparing approaches and translating the findings from one application to another.

@@

% Exact methods
Several articles propose exact methods. These are usually based on a Mixed Integer Linear Program (MILP) formulation that is solved with the help of a commercial solver \citep{bredstrom2008,bachouch2011, mankowska2014, aithaddadene2016}. Typically, these formulations are three index formulations that involve two types of variables: $x_{ijk}$ binary variables that indicate if carer $k$ traverses the arc $i$ and $j$; and $t_{ik}$ real variables that determine at what time carer $k$ starts performing job $i$. These formulations are limited in practice and can only solve to optimality small problems (e.g. 45 patients in \cite{aithaddadene2016} or 10 in \cite{mankowska2014}). In order to address this, some researchers propose to decompose the problem in a column generation scheme. For example, \cite{rasmussen2012} tackles the HHCRSP problem in Denmark. They propose a branch-and-price scheme where generalised synchronisation constraints are relaxed and dealt with in the branching. % \correction{Only Rasmussen?}
 % \correction{Cite relevant papers here}. The ILP formulations work well for small problems, but are often not practical to solve for problems with a significant number of patients. %around 50

Often, in order to tackle larger instances, the MILPs are solved in a restricted manner and, while they do not guarantee optimality, they might provide feasible solutions suitable for day to day planning. This is often referred to as a matheuristic algorithm and can be done by reducing the size of the original problem by fixing some aspects and solving it again. One example of this technique is given in \cite{rasmussen2012} that also proposes clustering strategies to reduce the size of the problem, which allows them to solve problems with up to $150$ patients, although no longer proving optimality. Another is \cite{bredstrom2008}, where they propose a heuristic that iteratively allows or forbids assignments between patients and carers in order to generate a series of restricted MILPs that can be solved in a short time.

Another approach, which is the one we use in this article, is to propose a metaheuristic algorithm independent of a mathematical solver. \cite{mankowska2014} and \cite{aithaddadene2016} propose an Adaptive Large Neighbourhood search and a hybrid GRASP combined with Iterated Local Search respectively. One advantage of the use of heuristic algorithms is that they do not require availability of a (often commercial) MILP solver, which is unlikely to be available to health care professionals.

Finally, we look at the literature that discusses the software implementation of the solution approaches in practice. One of the earliest works is due to \cite{begur1997}, who developed a decision support system based on the Clarke and Wright algorithm \citep{clarke1964} for vehicle routing and then relied on integration with a Geographic Information System (GIS), an interactive user interface and a human planner to alter the routes as necessary. The Laps Care system \citep{eveborn2006, eveborn2009} is a Swedish software for the HHCRSP, based on a repeated matching algorithm has been shown to be successful in saving planning time, minimising travel times and even reducing the number of missed visits. The authors estimated annual savings in 2009 of up to \euro 120 million across teams in Norway, Finland and Sweden \citep{eveborn2009}. Another software is Parpap \citep{bertels2006}, which proposes various approaches combining constraint programming and metaheuristics. A common claim in all these implementation-oriented works is that the software needs to be flexible enough to allow for real-world constraints to be implemented and offers interaction with mapping software.

%----------------------------------------------------------------------------------%
\section{A model for health home care services}
\label{sec:model}
\noindent In this section we describe the notation and give a formal description of the HHCRSP problem tackled by our DST.

We consider a list of nurses $N = \{1,\dots,n\}$ that are available for the day. Each nurse $i$ might have a different starting time, $st(i)$, and end time, $en(i)$, of their working day. During the planning horizon, which is one day in our case, the nurses need to perform a number of jobs $J = \{1,\dots,m\}$, a subset of which, $DS \subset J$ are double services, i.e. they require two nurses to be present to be performed. Note that jobs refer to individual tasks that a nurse needs to perform and are not necessarily identified with unique patients or locations, meaning that the same patient might have more than one job on the list.

@@

%----------------------------------------------------------------------------------%
\subsection{Skill mix}
\label{sub:skillmix}
\noindent Some jobs might require a specialised nurse to be performed, and nurses might only perform jobs that they are qualified to do. To model this, we define a function $S_1: N\times J \rightarrow \{0,1\}$, that takes value $1$ if nurse $i$ is skilled to perform job $j$, $S_1(i,j) = 1$ and $0$ otherwise. 
Naturally, for double service jobs, no single nurse is skilled to perform them $S_1(i,j) = 0, \forall i \in N, \forall j \in DS$. To account for this, we define a second function $S_2: N\times N\times J \rightarrow \{0,1\}$, that takes value $S_2(i_1,i_2,j) = 1$ if the combination of nurses $i_1$ and $i_2$ is sufficiently skilled to perform job $j$ and $0$ otherwise.

@@

A relevant modelling consideration relates to the necessary skill mix for simultaneous double services or, in other words, how the function $S_2$ is created in practice. We distinguish the following cases:
\begin{itemize}
    \item \textbf{Duplicated}. In this case it is required for all staff members visiting the patient to have all necessary skills to perform the service.
    \item \textbf{Shared}. This is the case when it is sufficient that the combination of skills between the staff members present is enough to perform the service. In particular, it could happen that only one staff member is qualified for all services and the other for none. An example of this would be visits that require a second person to provide support and ensure safety, for example with patients that have a record of behaving inappropriately towards staff.
    \item \textbf{Strictly shared}. In this case each member of the service does not need to be qualified for all services but should be attending the visit to perform at least one service. This is the approach used in \cite{mankowska2014} and \cite{aithaddadene2016}.
    \item \textbf{Shared-Duplicated}. This case is analogous to \emph{Shared}, but it allows some skills to be required by more than one of the staff members, hence it generalises the previous cases.
\end{itemize}

\noindent In our DST, the default approach is \emph{Shared-Duplicated}, as it allows for the flexibility that is observed in real operations. The data is introduced as a list of skills and, if some skills need to be present in more than one staff member, they can simply appear duplicated on the list. We have, however, also implemented the Strictly Shared approach, in order to solve the instances from the literature.

It is not uncommon in the literature to model the skills of the workforce as hierarchical, i.e. each nurse is assigned a skill level and it is assumed that they possess all the skills that lower levels have, and some more. The model might then introduce a penalty or limit for high skilled nurses performing too many low skilled jobs, known as downgrading \cite[see, for example,][]{fikar2015}. In the DN teams we observed, skills were not hierarchical, neither were any such limits in place. Nevertheless, nurses pointed out that high skilled nurses were typically assigned only the high skilled jobs, as they might need to perform other jobs (e.g. attending meetings or other administrative tasks). While this is not something that we consider explicitly, a similar objective may be achieved through setting negative preferences between high skilled nurses and low skilled jobs.

%----------------------------------------------------------------------------------%
\subsection{Preferences}
\label{sub:preferences}
\noindent It is not uncommon in practice that district nurses do their planning using preferences. These might arise from multiple reasons, such as a preserving continuity of care, gender preference, patients and nurses speaking the same language, previous conflicts between a patient and a staff member, \model{etc}. We model this with a preference function, $P: N\times J \rightarrow \mathbb{R}$, such that $P(i,j)$ represents the preference score of assigning job $i$ to staff member $j$. The function $P$ should assign a higher (positive) score to preferred patient-nurse assignments and a lower (negative) score to assignments that are better avoided. We give an indication on how to choose the specific values of the function in Section~\ref{subsub:adequateweights}.

%----------------------------------------------------------------------------------%
\subsection{Travel times}
\label{sub:traveltimes}
\noindent To describe travel times, for each nurse we define $N$ functions $T_i: J\cup\{k^i_0\} \times J\cup\{k^i_0\} \rightarrow \mathbb{R}$, $i \in N$, where $T_i(j_1, j_2)$ indicates the time it takes nurse $i$ to travel from job $j_1$ to job $j_2$, where $k^i_0$ refers to the start and end location of the nurse $i$. Each nurse might start from a different depot (typically, their own home), a fact that is accounted for in these functions. Note that with this definition we allow to have different travel times between jobs for different nurses, and this can be used to model, for example, different means of transport. In our observations, however, all nurses used their own vehicles, so we assume that $T_i$ returns driving times for all nurses and, in particular, that:
\begin{equation}
    T_{i_1}(j_1, j_2) = T_{i_2}(j_1, j_2),\ \forall i_1, i_2 \in N,\ \forall j_1, j_2 \in J. \label{eqn:traveltimequality}
\end{equation}

\noindent Note that this function can be substituted by a more sophisticated one without any changes to the algorithms described here.

@@

%----------------------------------------------------------------------------------%
\subsection{Temporal constraints}
\label{sub:temporalconstraints}
\noindent Each job $j$ has a service time $s_j$ and a time window $t_j$ that is defined by its earliest start time, $e_j$ and the latest start time $l_j$. It is a hard constraint that jobs cannot start before their earliest start time:
\begin{align}
    t_j \geq e_j, & \forall j \in J, \label{eqn:jobstart}
\end{align}

\noindent however, they may start late. The tardiness with which a job $j$ starts, $z_j$, can be simply calculated using the following:
\begin{equation}
    z_j = \max(0, t_j - l_j). \label{eqn:tardiness}
\end{equation}

\noindent Note that tardiness here refers only to the difference between the job start and its original time window. Jobs that have dependencies or require two nurses might be further restricted temporally. This might happen in practice, for example, in situations where a certain medication must be administered \model{a few hours (a specific time period/amount of time)} after the first dose. If two jobs $j_1$ and $j_2$ are dependent, this is modelled by imposing a second time window on $j_2$ that depends on the start time of $j_1$, $[e^{(j_1)}_{j_2} + t_{j_1}, l^{(j_1)}_{j_2} + t_{j_1}]$. Then, the same time window constraints hold:
\begin{subequations}
	\begin{align}
	t_{j_2} &\geq e^{(j_1)}_{j_2} + t_{j_1}, \label{eqn:timewindow1}\\[3pt] % & \forall j \in J
	t_{j_2} &\leq l^{(j_1)}_{j_2} + t_{j_1}. \label{eqn:timewindow2} % & \forall j \in J
	\end{align}
\end{subequations}

\noindent Similarly, $j_2$ will impose a time window on $j_1$, and in a valid instance of the problem it must hold that:
\begin{subequations}
	\begin{align}
	e^{(j_1)}_{j_2} &= -e^{(j_2)}_{j_1}, \label{eqn:earliesttimedependecies}\\[3pt]
	l^{(j_1)}_{j_2} &= -l^{(j_2)}_{j_1}. \label{eqn:latesttimedependencies}
	\end{align}
\end{subequations}

\noindent Then, we can define the \emph{gap tardiness} of job $j_2$, $\tilde{z}_{j_2}$ as:
\begin{equation}
    \tilde{z}_{j_2} = \max(0, t_{j_2} - (l^{(j_1)}_{j_2} + t_{j_1})). \label{eqn:gaptardiness}
\end{equation}

\noindent \model{Since some jobs have time windows (only \textit{some}?)} or might need to be synchronised, waiting times may occur. We denote by $w_{ij}$ the waiting time of nurse $i$ at job $j$. Note that $w_{ij} = 0$ if $j \notin r_i$; otherwise, $w_{ij}$ represents the difference between $t_j$ and the arrival time of nurse $i$ to job $j$.% , denoted by $a(i,j)$.

The arrival time $a(i,j)$ of nurse $i$ at job $j \in r_i$ is calculated as follows:
\begin{equation}
    a(i,j) = st(i) + \sum_{h=0}^{j' - 1} w_{ik^i_h} + s_{k^i_h} + T_i(k^i_h, k^i_{h + 1}),
\end{equation}

\noindent where $j'$ denotes the position of job $j$ in the route $r_i$, i.e. $j = k^i_{j'} \in r_i$. Note that there are no waiting or service times at the depot; that is, $w_{ik^i_0} = 0$ and $s_{k^i_0} = 0 $ for all $i \in N$.

In practice, waiting times may occur for a variety of reasons. In order to maintain simplicity in our model, we only consider three reasons for which a nurse $i$ might have waiting time $w_{ij} \geq 0$ at a job $j \in r_i$:
\begin{itemize}
    \item Respecting a time window: 
    \begin{align}
        w_{ij} \geq e_i - a(i,j) & \forall i \in N, \forall j \in J.
    \end{align}
    \item Waiting for a second nurse $i'$ to arrive for a double service:
    \begin{align}
        w_{i j} \geq t_j - a(i',j) & \forall j \in DS,\ such\ that\ j \in r_{i} \cap r_{i'}.
    \end{align}
    \item Waiting for the minimum gap of a synchronised service. If the job $j$ depends on $j'$, then:
    \begin{equation}\label{eqn:waitingforsynchronised}
        w_{ij} \geq e^{(j')}_{j} - a(i,j).
    \end{equation}
\end{itemize}
